Improving \textsl{Gaia} parallaxes with a data-driven model \\ of the color--magnitude diagram

### 2017-04-24 paperDistance/paper.tex
paper stuff

We correct for dust using 3D Bayestar (Schlafly+Finkbeiner) with an estimate of the distance from iteratively sampling our posterior distances. Dust corrections are non trivial for stars with poor signal to noise. Stars with poor signal to noise have a likelihood that is consistent with infinite distance and can therefore have severe dust corrections. This is most obvious in the giant stars, which tend to have the lowest signal to noise, and can therefore have dust corrections $> 2$ magnitudes. To apply a dust correction to the \tmass\ photometry, we first generate our prior using the observed, attenuated photometry. Using this raw prior we infer more precise distances to all the stars. We then use this more precise distance posterior to get a measurement of dust in the 3D dust map. This approach minimizes severe distances from the likelihoods alone and their associated severe dust corrections. In detail, we take the $5\%$ quantile of each distance posterior, so the closest bit of the posterior, and sample the Green et al 3D dust map at that distance. We take the $5\%$ quantile, as apposed to the median for example, to do a light dust correction because over correcting for dust can shoot the likelihood into an unphysical, different space of the CMD and then bias the distance inference. For each distance we sample the median of the probabilistic dust model. We apply each light dust correction to our \tmass\ photometry given by Equation \ref{eq:data} where $Q_{\lambda}$ = [0.709, 0.302] for bands [J, K] respectively, taken from table (REF TABLE SFD), and $E(B-V)$ is the output of the Green et al dust models. We then regenerate our prior using the lightly dust corrected photometry. We do not update the errors, but keep them the same as before the dust correction, which is wrong and under representing our errors but this leads again to a lighter deconvolution. We could add the covariance of the dust to our likelihoods, but this is beyond the scope of this demonstration of concept. We iterate this process 10 times, however the dust values seem to converge after about 2-3 iterations. The iterative process allows for slight variations in distance and corresponding dust to become more consistent with the \cmd. Within our probabilistic framework this handling of dust is not proper. We are taking a point estimate of the probabilistic dust model at a quantile distance of our posterior. A full, proper account would sample the dust model at distances sampled from the posterior and propagate those samples back into \xd during the iterative proccess. The issues is that this dust sample distribution would not be a Gaussian distribution which \xd\ requires. We could calculate this more complex dust distribution and estimate it as a Gaussian distribution to feed to \xd, however, this is beyond the scope of this paper.

@@

\begin{figure}
\centering
  \includegraphics[width=\textwidth]{posteriorCMD.png}
\caption{the expectation values and subsampled error bars}
\label{fig:posteriorCMD}
\end{figure}

@@

Some discussion of the inconsistency of how we treat dust within the probabilistic framework.

Some discussion of possible failure modes. Including very rare kinds of stars. Should we add gaussians to prior where we know stars exist that aren't well sampled in Gaia (ie WD)
Inference with our prior is bad for underrepresented populations in the \cmd. For example, it appears there are only $2$ white dwarfs in \gaia\ DR1. With t

Number of gaussians in \xd, and the w parameter. That we set them heuristically.

>>>>

We correct for dust using 3D Bayestar (Schlafly+Finkbeiner) with an estimate of the distance from iteratively sampling our posterior distances. Dust corrections are non trivial for stars with poor signal to noise. Stars with poor signal to noise have a likelihood that is consistent with infinite distance and can therefore have severe dust corrections. This is most obvious in the giant stars, which tend to have the lowest signal to noise, and can therefore have dust corrections $> 2$ magnitudes. To apply a dust correction to the \tmass\ photometry, we first generate our prior using the observed, attenuated photometry. Using this raw prior we infer more precise distances to all the stars. We then use this more precise distance posterior to get a measurement of dust in the 3D dust map. This approach minimizes severe distances from the likelihoods alone and their associated severe dust corrections. In detail, we take the $5\%$ quantile of each distance posterior, so the closest bit of the posterior, and sample the Green et al 3D dust map at that distance. We take the $5\%$ quantile, as apposed to the median for example, to do a light dust correction because over correcting for dust can shoot the likelihood into an unphysical, different space of the CMD and then bias the distance inference. For each distance we sample the median of the probabilistic dust model. We apply each light dust correction to our \tmass\ photometry given by Equation \ref{eq:data} where $Q_{\lambda}$ = [0.709, 0.302] for bands [J, K] respectively, taken from table (REF TABLE SFD), and $E(B-V)$ is the output of the Green et al dust models. We then regenerate our prior using the lightly dust corrected photometry. We do not update the errors, but keep them the same as before the dust correction, which is wrong and under representing our errors but this leads again to a lighter deconvolution. We could add the covariance of the dust to our likelihoods, but this is beyond the scope of this demonstration of concept. We iterate this process 10 times, however the dust values seem to converge after about 2-3 iterations. The iterative process allows for slight variations in distance and corresponding dust to become more consistent with the \cmd.

@@

\begin{figure}
\centering
  \includegraphics[width=\textwidth]{posteriorCMD.png}
\caption{the expectation values and subsampled error bars. Notice that some parts of this \cmd\ are tighter than the prior, or noise-deconvolved \cmd. The visual of the prior is a sampling of the Gaussians, but this is the expectation value so this lies much closer to the center of the distribution. Similar to the toy model, where the pdf maximum of the noisier points lies in the center of the distribution.}
\label{fig:posteriorCMD}
\end{figure}

@@

\subsection{Proper Dust Corrections}
Within our probabilistic framework, the way we handle dust is not proper. We are taking a point estimate of the probabilistic dust model at a quantile distance of our posterior, correcting our photometry using this point estimate and rederiving the prior in \xd. A full, proper account would sample the dust model at distances sampled from the posterior and propagate those samples back into \xd during the iterative proccess. The issues is that this dust sample distribution would not be a Gaussian distribution which \xd\ requires. We could calculate this more complex dust distribution and estimate it as a Gaussian distribution to feed to \xd, however, this is beyond the scope of this paper.

\subsection{Stationarity Assumption}
Both the low and high signal to noise objects are being drawn from the same distribution, independent of noise. This isn't precisely true because most Gaia objects have similar noise but their signal depends on their distance. Nearby stars have higher signal (parallax) because they are closer and the nearby stars are mostly main sequence stars (IMF, timing, malmquist). We can see bright giants to much larger distances and therefore lower signal or parallax and lower signal to noise. So the red giant branch is a population of relatively lower signal to noise objects than the main sequence. But I think this is ok, that the low and high signal to noise objects have to be drawn from the same distribution within some feature or patch of the space. This is all assuming the Gaia noise model is correct. If the uncertainties are over estimated then we could be over deconvolving (generating a tighter distribution than the truth) And in some extreme over estimate of uncertainties the lower signal to noise giants could be pulled to the main sequence. But with the quoted uncertainties this seems unlikely. As you move up the giant branch and probe farther giants the signal to noise will continue to diminish but I think this is okay since the color uncertainties are not large. Possibly a metallicity trend with distance, placing more nearby (higher S/N) objects at slightly different positions on the CMD pulling lower S/N objects to that distribution for our prior. But if something has small error bars is good data is good data both for inference and generating the prior

For very rare kinds of stars. Inference with our prior is bad for underrepresented populations in the \cmd. For example, it appears there are only $2$ white dwarfs in \gaia\ DR1. With them being very underrepresented in the prior, the amplitudes of the gaussians representing them is very small, so the WDs are more likely to be pulled to mimic a much farther away MS star.

\subsection{Cross Validation}
For generating our prior, we use a mixture of 128 Gaussians. This number, 128, was chosen heuristically. It makes the method computationally tractable, and is sufficient for this demonstration of concept, but more Gaussians might be desired for a more final data product. We would ideally use the optimal number of Gaussians determined by cross validation, but preliminary tests using cross validation suggest the optimal number of Gaussians is very large.
